---
title: "How do purpose in life and happiness differ?"
subtitle: "Correlates and consequences over two years"
author: 
  - name: Todd B. Kashdan
    orcid: 0000-0001-6438-0485
    email: tkashdan@gmu.edu
    affiliation: 
      - ref: GMU
    equal-contributor: true
    corresponding: true
  - name: Patrick E. McKnight
    orcid: 0000-0002-9067-9066
    email: pmcknigh@gmu.edu
    affiliation: 
      name: George Mason University
      department: Department of Psychology
      id: GMU
    equal-contributor: true
    corresponding: false
  - name: Graduate Student
    email: student@gmu.edu
    affiliation:
      - ref: GMU
  - name: Madeleine Gross
    affiliation:
      name: University of California, Santa Barbara
      department: Department of Psychology
author-note: 
  disclosures:
    data-sharing: "The data that support the findings of this study are available from the corresponding author, PEM, upon reasonable request."	
    authorship-agreements: "The first two authors contributed equally to this work; we determined the order quasi-randomly with replacement and laughter."
format:
  apaquarto-html: default
  apaquarto-docx: default
  apaquarto-pdf: default

bibliography: references.bib
abstract: "In the social sciences, a case has been made for distinguishing between purpose in life and happiness. That said, a surprisingly small number of direct empirical comparisons exist. Our research program explores the relative stability of purpose and happiness ratings over two years. In addition, we explored whether purpose and happiness change independently of each other over time. Furthermore, we tested what best differentiates purpose and happiness using a comprehensive battery of personality constructs (e.g., goal-specific hope scale, psychological needs, distress tolerance, and values). A community sample of XXX participants (mean age = XX, *SD* = XX; XX% women, XX% European-American) completed purpose in life and happiness measures at baseline and 6-month and 2-year intervals. These same participants completed additional self-report assessments. Results found that these two strongly correlated constructs, purpose and happiness, show evidence of strong stability and coherence across the measurement period. Predictors of change in purpose and happiness differed meaningfully. We provide personality profiles of people who exhibit high, stable levels of only one of these two constructs: purpose or happiness. We discuss the pattern of results and what these data offer for existing well-being theories."
keywords: 
  - purpose in life
  - happiness
  - correlates
  - stability
  - measurement
  - values
  - personality
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| error: false
#| warning: false
#| include: false

# Load necessary libraries
library(lavaan)       # For CFA analysis
library(dplyr)        # For data manipulation
library(tidyr)        # For data reshaping
library(psych)        # For descriptive statistics
library(ggplot2)      # For visualization
library(gridExtra)    # For arranging plots
library(knitr)        # For table formatting
library(kableExtra)   # For enhanced tables
library(flextable)    # For enhanced tables
library(stringr)      # For string manipulation
library(foreign)      # For reading SPSS files
library(corrplot)     # For correlation plots
library(car)          # For regression analysis
library(broom)        # For tidy model output
library(gt)           # For publication tables
library(tidyverse)    # For comprehensive data manipulation
library(haven)
```

# Introduction

There are multiple paths to a worthwhile existence. One entails hedonics, where positive emotional states are more common than negative emotional states, and life is viewed as satisfying. Another involves purpose, where life decisions and goals are centrally influenced by one's ultimate concerns (not the mere attainment of pleasure and avoidance of suffering; Bronk, 2014; McKnight & Kashdan, 2009). Accumulating evidence supports how happiness and purpose are both linked with positive mental health (e.g., Diener & Chan, 2011; Pfund & Hill, 2018). While the same can be said for many positively valenced constructs such as optimism (Peterson, 2000), kindness (Hui, Ng, Berzaghi, Cunningham-Amos, & Kogan, 2020), and mindfulness (Brown, Ryan & Creswell, 2007), happiness and purpose are grander than desirable personality traits (Disabato, Goodman, & Kashdan, 2025). Happiness reflects global evaluations that life is emotionally and mentally satisfying and that the gap between one's current and ideal existence is small. Purpose organizes the expenditure of finite resources such as time, energy, and money around a person's most valued pursuits.

## Proposed Distinctions Between Happiness and Purpose

To better understand happiness and purpose, we examined where the two might meaningfully differ in their benefits. Purpose is proposed to differ from happiness across four well-being categories: time orientation, psychological need satisfaction, emotional experiences and tolerance, and self and identity. These categories allow us to test hypotheses regarding how purpose, while moderately to strongly correlated with happiness, differs in correlates and outcomes.

### Time Orientation

Purpose is hypothesized to be more strongly linked to constructs that emphasize long-term striving and goal pursuit, whereas happiness is more closely tied to present-moment experiences. Hope, as conceptualized by psychologists (Snyder et al., 1991), exists when a person believes goals are achievable - recognizing *pathways* to making progress and *agency* to use those pathways. Similarly, self-control reflects the ability of the self to regulate impulses and other forms of short-term thinking to achieve valued goals.

### Psychological Need Satisfaction

Self-determination theory (SDT) describes the fulfillment of three basic needs as critical to living well: autonomy, belonging, and competence (Deci & Ryan, 2000). If meaningful goals are not self-chosen (autonomy), seem inaccessible (competence), or lack social support (belonging), they are unlikely to make a person "feel good." Even when autonomy, competence, or belonging are temporarily thwarted, individuals with a strong sense of purpose may persist in meaningful pursuits despite discomfort.

### Emotional Experiences and Tolerance

Purpose is also hypothesized to have a unique relationship with stressful events and resulting distress. Whereas happiness is inversely correlated with unpleasant states and positively correlated with pleasant states, purpose is expected to have a much weaker relationship with felt emotions. Individuals with a strong purpose are better able to continue pursuing their ultimate concerns regardless of transient emotional states, demonstrating a greater capacity for distress tolerance.

### Self and Identity

Unlike happiness, which is closely tied to pleasure and enjoyment, purpose is hypothesized to be more strongly associated with values that transcend momentary pleasure. Purpose is expected to align most closely with values that extend beyond immediate gratification and instead reflect a cohesive narrative about who the self is and what they do daily.

## The Present Research

In this research program, we examined how happiness and purpose differ in their correlates and consequences. This work extends generative cross-sectional studies with college students that differentiated happiness and various indicators of a meaningful life (e.g., Baumeister et al., 2013; Diener, Fujita, Tay, & Biswas-Diener, 2012). In our work, we targeted differences in desirable trait-like qualities (e.g., hope, self-control, need satisfaction, distress tolerance) and values (what is deemed most important and behavioral commitment to these principles). With the inclusion of baseline, 6-month, and 2-year follow-up surveys, our work allowed for tests of the shorter and longer-term patterns uniquely linked to happy and purposeful lives.

# Method

## Participants and Procedure

Community adults were recruited from the DC/Maryland/Virginia region through local advertisements. The baseline sample (Time one; T1 *n* = 303) completed trait measures and ideographic assessments in the laboratory and subsequently completed follow-up measures six months (Time two; T2 *n* = 205) and two years later (Time three; T3 *n* = 167) through an online survey platform.

## Measures

### Purpose

**Brief Measure of Purpose in Life (BPIL; Hill, Edmonds, Peterson, Luyckx, & Andrews, 2016).** The 4-item BPIL measures the degree to which one has a clear mission in life (e.g., *"My plans for the future match with my true interests and values"*). Items are rated on a 5-point Likert scale (from 0 or *not at all* to 4 or *very much*). The BPIL demonstrates good construct validity and exhibited acceptable internal consistency at all time points (T1: α = .84; T2: α = .90; T3: α = .88).

We also used the **Purpose in Life - External Aim Scale (PILEA; Hill, Edmonds, Peterson, Luyckx, & Andrews, 2016)**. The PILEA is a 6-item measure that assesses purpose in life through two subscales: engagement (e.g., *"I feel that I am doing what I was meant to do in life"*) and beyond-the-self (e.g., *"I feel that my life has meaning beyond my own personal interests"*). Items are rated on a 5-point Likert scale (from 0 or *not at all* to 4 or *very much*). The PILEA also demonstrated good construct validity and exhibited acceptable internal consistency at all time points (T1: α = .90; T2: α = .92; T3: α = .91).

### Happiness

**Subjective Happiness Scale (SHS; Lyubomirsky & Lepper, 1999).** The 4-item SHS assesses global subjective happiness using a 7-point Likert scale with different anchors based on individual items. The SHS demonstrates acceptable internal consistency in the present study (T1: α = .87; T2: α = .89; T3: α = .89).

### Correlate Measures

**Goal Specific Hope Scale (GSHS; Feldman, Rand, & Kahle-Wrobleski, 2009).** Measures hope related to accomplishing a goal with agency and pathways subscales.

**Self-Control Scale (SCS; Tangney, Baumeister, & Boone, 2004).** Measures the capacity to adapt the self to achieve a better fit with the environment.

**Balanced Measure of Psychological Needs (BMPNS; Sheldon & Hilpert, 2012).** Measures satisfaction of basic needs: autonomy, competence, and belonging.

**Distress Intolerance (McHugh & Otto, 2012).** Measures one's inability to tolerate negative somatic and emotional states.

**Brief Experiential Avoidance Measure (BEAQ; Gámez et al., 2014).** Measures experiential avoidance as a shortened version of the MEAQ.

**Values Measures.** The Twenty Item Values Inventory assesses both importance of particular values and behavioral consistency with those values across 10 value domains.

**Stress Tolerance.** A 6-item measure of stress tolerance was created for this study, assessing the ability to cope with stressors and maintain functioning.

## Data Loading and Preparation


```{r}
#| label: data-loading
#| echo: false
#| message: false
#| warning: false

# Read the data
#dat <- read_sav("Psychological Flexibility_Full Dataset_070822.sav")
dat <- read.csv("alldatPvH4LLM.csv")[, -1]

# Define column ranges for key variables based on the specification document
# Purpose and Happiness variables
purp_hap_cols <- c(
  # ID variable
  1,
  # Purpose (BPURP) - baseline, FU1, FU2  
  398:401, 697, 735,  # baseline items and means
  1467:1470, 1771, 1845,  # FU1 items and means
  3001:3004, 3542, 3556,  # FU2 items and means
  
  # Purpose in Life - External Aim Scale (PILEA) - baseline, FU1, FU2
  402:407, 698, 1888, 1889, # baseline items and means
  1471:1476, 1772, 1846, 1892, 1893, # FU1 items and means & POMP
  3005:3010, 3543, 3544, 3545, # FU2 items and means
  
  # Happiness (SHS) - baseline, FU1, FU2
  241:243, 245, 673, 730, # baseline:  excludes non-reverse scored items, mean & POMP scores
  1419:1421, 1423, 1751, 1825,  # FU1 items and means 
  2954:2956, 3428, 3466, 3555  # FU2 items and means
)

# Extract basic purpose and happiness data
basic_dat <- dat[, purp_hap_cols]

# Define comprehensive variable extraction based on document specifications
comprehensive_cols <- c(
  # ID
  1,
  
  # TEMPORAL FOCUS - Goal Specific Hope Scale (GSHS)
  # Baseline: columns 447, 448-451, 449-453, 709-711 (adjusted for 0-indexing)
  441, 442, 444, 442, 445, 446, 702:704, 702:704, # NOTE: no POMP scores - divide means by 7 for the second set of 702:704 so that it matches the rest
  # FU1: 1477, 1479, 1480 path, 1478, 1481, 1482 agen items; 1773 mean, 1774 path mean, 1775 agen mean
  1477, 1479, 1480, 1478, 1481, 1482, 1773:1775, 1847:1849,
  # FU2: multiple goal items and means
  3011, 3013, 3014, 3012, 3015, 3016, 3479:3481, 3479:3481, # NOTE no POMP scores so divide means by 7 for the second set of 3479:3481
  
  # Self-control (SCS) 
  # Baseline: 480:489 (raw items I omitted); 709:715 reverse scored items; 716 mean
  709:716,
  
  # NEEDS - Basic Psychological Needs Scale (BMPNS)
  # Baseline items and means
  246:248, 264:266, 252:254, 267:269, 258:260, 270:272, 674:685, 731:733,
  # FU1 items and means  
  1424:1426, 1739:1741, 1430:1432, 1742:1744, 1436:1438, 1745:1750, 1822:1824,
  # FU2 items and means
  2958:2960, 3434:3436, 2964:2966, 3437:3439, 2970:2972, 3440:3442, 3492:3503,
  
  # EMOTIONS - Distress intolerance, experiential avoidance, etc.
  430:439, 701, # Distress intolerance baseline
  408:412, 627, 414:422, 699, 738,  # Experiential avoidance 
  468:469, 628:629, 705,  # Implicit beliefs about emotions
  
  # VALUES - Importance and behavioral consistency
  278:297, 687:696,  # Baseline importance no POMP scores
  526:545, 637:646,  # Baseline behavioral consistency no POMP scores
  1447:1466, 1753:1762, 1827:1836,  # FU1 importance with POMP scores
  1483:1502,  # FU1 behavioral consistency with no means and no POMP scores
  2981:3000, 3504:3513,  # FU2 importance
  3023:3042, 3514:3523,   # FU2 behavioral consistency
  
  # Stress Tolerance
  185:190, # baseline stress tolerance items with mean
  1899:1904, # FU1 stress tolerance items with mean
  3429:3433, 3472 # FU2 stress tolerance items with mean
  
)

# Extract comprehensive dataset (adjust for R's 1-based indexing)
comprehensive_dat <- dat[, comprehensive_cols]
```

### Pomp Conversion

```{r}
#| label: pomp-conversion
#| echo: false
#| message: false
#| warning: false

# Function to convert to POMP scores (0-100 scale)
convert_to_pomp <- function(x, min_val, max_val) {
  ((x - min_val) / (max_val - min_val)) * 100
}

# Convert key measures to POMP scores
# Purpose (BPURP): 0-4 scale
basic_dat$b_purpose_pomp <- convert_to_pomp(basic_dat$b_bpurp_mean, 0, 4)
basic_dat$fu1_purpose_pomp <- convert_to_pomp(basic_dat$fu1_bpurp_mean, 0, 4)  
basic_dat$fu2_purpose_pomp <- convert_to_pomp(basic_dat$fu2_bpurp_mean, 0, 4)

# Purpose in Life - External Aim Scale (PILEA): 0-4 scale
basic_dat$b_pilea_pomp <- convert_to_pomp(basic_dat$b_pilea_mean, 0, 4)
basic_dat$b_pilea_eng_pomp <- convert_to_pomp(basic_dat$b_pilea_eng_mean, 0, 4)
basic_dat$b_pilea_bey_pomp <- convert_to_pomp(basic_dat$b_pilea_bey_mean, 0, 4)
basic_dat$fu1_pilea_pomp <- convert_to_pomp(basic_dat$fu1_pilea_mean, 0, 4)
basic_dat$fu1_pilea_eng_pomp <- convert_to_pomp(basic_dat$fu1_pilea_eng_mean, 0, 4)
basic_dat$fu1_pilea_bey_pomp <- convert_to_pomp(basic_dat$fu1_pilea_bey_mean, 0, 4)
basic_dat$fu2_pilea_pomp <- convert_to_pomp(basic_dat$fu2_pilea_mean, 0, 4)
basic_dat$fu2_pilea_eng_pomp <- convert_to_pomp(basic_dat$fu2_pilea_eng_mean, 0, 4)
basic_dat$fu2_pilea_bey_pomp <- convert_to_pomp(basic_dat$fu2_pilea_bey_mean, 0, 4)

# Happiness (SHS): 0-6 scale  
basic_dat$b_happiness_pomp <- convert_to_pomp(basic_dat$b_shs_mean, 0, 6)
basic_dat$fu1_happiness_pomp <- convert_to_pomp(basic_dat[[30]], 0, 6)
basic_dat$fu2_happiness_pomp <- convert_to_pomp(basic_dat[[36]], 0, 6)

# select only the variables with mean in the name from comprehensive_dat
comprehensive_dat_means <- comprehensive_dat %>%
  select(1, contains("mean"), contains("tot"), contains("fu1_valc"))

# convert GSHS to POMP scores in comprehensvive_dat_means
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    b_gshs_mean_pomp = convert_to_pomp(b_gshs_mean, 0, 7),
    b_gshs_path_pomp = convert_to_pomp(b_gshs_path_mean, 0, 7),
    b_gshs_agen_pomp = convert_to_pomp(b_gshs__agen_mean, 0, 7),
    fu1_gshs_mean_pomp = convert_to_pomp(fu1_gshs_mean, 0, 7),
    fu1_gshs_path_pomp = convert_to_pomp(fu1_gshs_path_mean, 0, 7),
    fu1_gshs_agen_pomp = convert_to_pomp(fu1_gshs_agen_mean, 0, 7),
    fu2_gshs_mean_pomp = convert_to_pomp(fu2_gshs1_mean, 0, 7),
    fu2_gshs_path_pomp = convert_to_pomp(fu2_gshs1_path_mean, 0, 7),
    fu2_gshs_agen_pomp = convert_to_pomp(fu2_gshs1_agen_mean, 0, 7)
  )

# convert SCS to POMP scores in comprehensive_dat_means
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    b_self_control_pomp = convert_to_pomp(b_scs_mean, 0, 3) # self-control
  )
# convert BMPNS to POMP scores in comprehensive_dat_means
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    b_autonomy_pomp = convert_to_pomp(b_bmpns_A_tot, 0, 4), # autonomy
    b_competence_pomp = convert_to_pomp(b_bmpns_C_tot, 0, 4), # competence
    b_relatedness_pomp = convert_to_pomp(b_bmpns_R_tot, 0, 4), # relatedness
    fu1_autonomy_pomp = convert_to_pomp(fu1_bmpns_A_tot, 0, 4),
    fu1_competence_pomp = convert_to_pomp(fu1_bmpns_C_tot, 0, 4),
    fu1_relatedness_pomp = convert_to_pomp(fu1_bmpns_R_tot, 0, 4),
    fu2_autonomy_pomp = convert_to_pomp(fu2_bmpns_A_tot, 0, 4),
    fu2_competence_pomp = convert_to_pomp(fu2_bmpns_C_tot, 0, 4),
    fu2_relatedness_pomp = convert_to_pomp(fu2_bmpns_R_tot, 0, 4)
  )

# convert Distress Intolerance to POMP scores in comprehensive_dat_means
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    b_distress_intolerance_pomp = convert_to_pomp(b_di_mean, 0, 5)
  )

# convert experiential avoidance (BMEAQ) and implicit beliefs about emotions to POMP scores in comprehensive_dat_means
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    b_experiential_avoidance_pomp = convert_to_pomp(b_bmeaq_mean, 0, 6),
    b_implicit_beliefs_emotions_pomp = convert_to_pomp(b_ibaes_mean, 0, 6)
  )

# compute means for fu1_valc 
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    fu1_valc_con_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_con")), na.rm = TRUE),
    fu1_valc_trad_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_trad")), na.rm = TRUE),
    fu1_valc_ben_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_ben")), na.rm = TRUE),
    fu1_valc_univ_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_univ")), na.rm = TRUE),
    fu1_valc_sd_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_sd")), na.rm = TRUE),
    fu1_valc_stim_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_stim")), na.rm = TRUE),
    fu1_valc_hed_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_hed")), na.rm = TRUE),
    fu1_valc_ach_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_ach")), na.rm = TRUE),
    fu1_valc_pow_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_pow")), na.rm = TRUE),
    fu1_valc_sec_mean = rowMeans(select(comprehensive_dat_means, starts_with("fu1_valc_sec")), na.rm = TRUE)
  )

# convert Values Behavioral Consistency to POMP scores in comprehensive_dat_means
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    b_vali_con_pomp = convert_to_pomp(b_vali_con_mean, 0, 6),
    b_vali_trad_pomp = convert_to_pomp(b_vali_trad_mean, 0, 6),
    b_vali_ben_pomp = convert_to_pomp(b_vali_ben_mean, 0, 6),
    b_vali_univ_pomp = convert_to_pomp(b_vali_univ_mean, 0, 6),
    b_vali_sd_pomp = convert_to_pomp(b_vali_sd_mean, 0, 6),
    b_vali_stim_pomp = convert_to_pomp(b_vali_stim_mean, 0, 6),
    b_vali_hed_pomp = convert_to_pomp(b_vali_hed_mean, 0, 6),
    b_vali_ach_pomp = convert_to_pomp(b_vali_ach_mean, 0, 6),
    b_vali_pow_pomp = convert_to_pomp(b_vali_pow_mean, 0, 6),
    b_vali_sec_pomp = convert_to_pomp(b_vali_sec_mean, 0, 6),
    
    b_valc_con_pomp = convert_to_pomp(b_valc_con_mean, 0, 6),
    b_valc_trad_pomp = convert_to_pomp(b_valc_trad_mean, 0, 6),
    b_valc_ben_pomp = convert_to_pomp(b_valc_ben_mean, 0, 6),
    b_valc_univ_pomp = convert_to_pomp(b_valc_univ_mean, 0, 6),
    b_valc_sd_pomp = convert_to_pomp(b_valc_sd_mean, 0, 6),
    b_valc_stim_pomp = convert_to_pomp(b_valc_stim_mean, 0, 6),
    b_valc_hed_pomp = convert_to_pomp(b_valc_hed_mean, 0, 6),
    b_valc_ach_pomp = convert_to_pomp(b_valc_ach_mean, 0, 6),
    b_valc_pow_pomp = convert_to_pomp(b_valc_pow_mean, 0, 6),
    b_valc_sec_pomp = convert_to_pomp(b_valc_sec_mean, 0, 6),
    
    fu1_vali_con_pomp = convert_to_pomp(fu1_vali_con_mean, 0, 6),
    fu1_vali_trad_pomp = convert_to_pomp(fu1_vali_trad_mean, 0, 6),
    fu1_vali_ben_pomp = convert_to_pomp(fu1_vali_ben_mean, 0, 6),
    fu1_vali_univ_pomp = convert_to_pomp(fu1_vali_univ_mean, 0, 6),
    fu1_vali_sd_pomp = convert_to_pomp(fu1_vali_sd_mean, 0, 6),
    fu1_vali_stim_pomp = convert_to_pomp(fu1_vali_stim_mean, 0, 6),
    fu1_vali_hed_pomp = convert_to_pomp(fu1_vali_hed_mean, 0, 6),
    fu1_vali_ach_pomp = convert_to_pomp(fu1_vali_ach_mean, 0, 6),
    fu1_vali_pow_pomp = convert_to_pomp(fu1_vali_pow_mean, 0, 6),
    fu1_vali_sec_pomp = convert_to_pomp(fu1_vali_sec_mean, 0, 6),
    
    fu1_valc_con_pomp = convert_to_pomp(fu1_valc_con_mean, 0, 6),
    fu1_valc_trad_pomp = convert_to_pomp(fu1_valc_trad_mean, 0, 6),
    fu1_valc_ben_pomp = convert_to_pomp(fu1_valc_ben_mean, 0, 6),
    fu1_valc_univ_pomp = convert_to_pomp(fu1_valc_univ_mean, 0, 6),
    fu1_valc_sd_pomp = convert_to_pomp(fu1_valc_sd_mean, 0, 6),
    fu1_valc_stim_pomp = convert_to_pomp(fu1_valc_stim_mean, 0, 6),
    fu1_valc_hed_pomp = convert_to_pomp(fu1_valc_hed_mean, 0, 6),
    fu1_valc_ach_pomp = convert_to_pomp(fu1_valc_ach_mean, 0, 6),
    fu1_valc_pow_pomp = convert_to_pomp(fu1_valc_pow_mean, 0, 6),
    fu1_valc_sec_pomp = convert_to_pomp(fu1_valc_sec_mean, 0, 6),
    
    fu2_vali_con_pomp = convert_to_pomp(fu2_vali_con_mean, 0, 6),
    fu2_vali_trad_pomp = convert_to_pomp(fu2_vali_tra_mean, 0, 6),
    fu2_vali_ben_pomp = convert_to_pomp(fu2_vali_ben_mean, 0, 6),
    fu2_vali_univ_pomp = convert_to_pomp(fu2_vali_uni_mean, 0, 6),
    fu2_vali_sd_pomp = convert_to_pomp(fu2_vali_sd_mean, 0, 6),
    fu2_vali_stim_pomp = convert_to_pomp(fu2_vali_stim_mean, 0, 6),
    fu2_vali_hed_pomp = convert_to_pomp(fu2_vali_hed_mean, 0, 6),
    fu2_vali_ach_pomp = convert_to_pomp(fu2_vali_ach_mean, 0, 6),
    fu2_vali_pow_pomp = convert_to_pomp(fu2_vali_pow_mean, 0, 6),
    fu2_vali_sec_pomp = convert_to_pomp(fu2_vali_sec_mean, 0, 6), 
    
    fu2_valc_con_pomp = convert_to_pomp(fu2_valc_con_mean, 0, 6),
    fu2_valc_trad_pomp = convert_to_pomp(fu2_valc_tra_mean, 0, 6),
    fu2_valc_ben_pomp = convert_to_pomp(fu2_valc_ben_mean, 0, 6),
    fu2_valc_univ_pomp = convert_to_pomp(fu2_valc_uni_mean, 0, 6),
    fu2_valc_sd_pomp = convert_to_pomp(fu2_valc_sd_mean, 0, 6),
    fu2_valc_stim_pomp = convert_to_pomp(fu2_valc_stim_mean, 0, 6),
    fu2_valc_hed_pomp = convert_to_pomp(fu2_valc_hed_mean, 0, 6),
    fu2_valc_ach_pomp = convert_to_pomp(fu2_valc_ach_mean, 0, 6),
    fu2_valc_pow_pomp = convert_to_pomp(fu2_valc_pow_mean, 0, 6),
    fu2_valc_sec_pomp = convert_to_pomp(fu2_valc_sec_mean, 0, 6)
)
  

# convert Stress Tolerance to POMP scores in comprehensive_dat_means
comprehensive_dat_means <- comprehensive_dat_means %>%
  mutate(
    b_stress_tolerance_pomp = convert_to_pomp(b_5dc_st_mean, 0, 6),
    fu1_stress_tolerance_pomp = convert_to_pomp(fu1_5dc_st_mean, 0, 6),
    fu2_stress_tolerance_pomp = convert_to_pomp(fu2_5dc_st_mean, 0, 6)
  )

# Select only the POMP converted columns for further analysis
comprehensive_dat_pomp <- comprehensive_dat_means %>%
  select(
    1, # ID
    contains("pomp") # Select all columns with "pomp" in their name
  )

datFinal <- basic_dat %>%
  left_join(comprehensive_dat_pomp, by = "id") %>%
  select(-contains("mean"), -contains("tot"), contains("pomp")) # Remove mean and total columns to avoid redundancy

# Save the final dataset for further analysis
write.csv(datFinal, "datFinal_pomp.csv", row.names = FALSE)

# create a new dataset (by Claude) for correlation analysis

correlation_dat <- dat %>%
  mutate(
    # Purpose measures (0-4 scale converted to POMP)
    purpose_bpurp_b = convert_to_pomp(b_bpurp_mean, 0, 4),
    purpose_bpurp_fu1 = convert_to_pomp(fu1_bpurp_mean, 0, 4),
    purpose_bpurp_fu2 = convert_to_pomp(fu2_bpurp_mean, 0, 4),
    
    purpose_pilea_total_b = convert_to_pomp(b_pilea_mean, 0, 4),
    purpose_pilea_total_fu1 = convert_to_pomp(fu1_pilea_mean, 0, 4),
    purpose_pilea_total_fu2 = convert_to_pomp(fu2_pilea_mean, 0, 4),
    
    purpose_pilea_eng_b = convert_to_pomp(b_pilea_eng_mean, 0, 4),
    purpose_pilea_eng_fu1 = convert_to_pomp(fu1_pilea_eng_mean, 0, 4),
    purpose_pilea_eng_fu2 = convert_to_pomp(fu2_pilea_eng_mean, 0, 4),
    
    purpose_pilea_bey_b = convert_to_pomp(b_pilea_bey_mean, 0, 4),
    purpose_pilea_bey_fu1 = convert_to_pomp(fu1_pilea_bey_mean, 0, 4),
    purpose_pilea_bey_fu2 = convert_to_pomp(fu2_pilea_bey_mean, 0, 4),
    
    # Happiness (1-7 scale converted to POMP)
    happiness_b = convert_to_pomp(b_shs_mean, 1, 7),
    happiness_fu1 = convert_to_pomp(fu1_shs_mean, 1, 7),
    happiness_fu2 = convert_to_pomp(fu2_shs_mean, 1, 7),
    
    # Time Orientation measures
    hope_total_b = convert_to_pomp(b_gshs_mean, 1, 8),
    hope_total_fu1 = convert_to_pomp(fu1_gshs_mean, 1, 8),
    hope_total_fu2 = convert_to_pomp(fu2_gshs_mean, 1, 8),
    
    hope_pathways_b = convert_to_pomp(b_gshs_path_mean, 1, 8),
    hope_pathways_fu1 = convert_to_pomp(fu1_gshs_path_mean, 1, 8),
    hope_pathways_fu2 = convert_to_pomp(fu2_gshs_path_mean, 1, 8),
    
    hope_agency_b = convert_to_pomp(b_gshs__agen_mean, 1, 8),
    hope_agency_fu1 = convert_to_pomp(fu1_gshs_agen_mean, 1, 8),
    hope_agency_fu2 = convert_to_pomp(fu2_gshs_agen_mean, 1, 8),
    
    # Self-control (1-4 scale)
    self_control_b = convert_to_pomp(b_scs_mean, 1, 4),
    
    # Psychological Needs (1-5 scale converted to POMP)
    autonomy_b = convert_to_pomp(b_bmpns_A_tot, 1, 5),
    autonomy_fu1 = convert_to_pomp(fu1_bmpns_A_tot, 1, 5),
    autonomy_fu2 = convert_to_pomp(fu2_bmpns_A_tot, 1, 5),
    
    competence_b = convert_to_pomp(b_bmpns_C_tot, 1, 5),
    competence_fu1 = convert_to_pomp(fu1_bmpns_C_tot, 1, 5),
    competence_fu2 = convert_to_pomp(fu2_bmpns_C_tot, 1, 5),
    
    relatedness_b = convert_to_pomp(b_bmpns_R_tot, 1, 5),
    relatedness_fu1 = convert_to_pomp(fu1_bmpns_R_tot, 1, 5),
    relatedness_fu2 = convert_to_pomp(fu2_bmpns_R_tot, 1, 5),
    
    # Emotional Tolerance measures
    distress_intolerance_b = convert_to_pomp(b_di_mean, 1, 5),
    experiential_avoidance_b = convert_to_pomp(b_bmeaq_mean, 1, 6),
    stress_tolerance_b = convert_to_pomp(b_5dc_st_mean, 1, 6),
    stress_tolerance_fu1 = convert_to_pomp(fu1_5dc_st_mean, 1, 6),
    stress_tolerance_fu2 = convert_to_pomp(fu2_5dc_st_mean, 1, 6),
    
    # Values measures (1-7 scale converted to POMP)
    # Self-transcendent values (beyond the self)
    values_benevolence_imp_b = convert_to_pomp(b_vali_ben_mean, 1, 7),
    values_universalism_imp_b = convert_to_pomp(b_vali_univ_mean, 1, 7),
    values_benevolence_cons_b = convert_to_pomp(b_valc_ben_mean, 1, 7),
    values_universalism_cons_b = convert_to_pomp(b_valc_univ_mean, 1, 7),
    
    # Self-enhancement values (personal focus)  
    values_achievement_imp_b = convert_to_pomp(b_vali_ach_mean, 1, 7),
    values_hedonism_imp_b = convert_to_pomp(b_vali_hed_mean, 1, 7),
    values_achievement_cons_b = convert_to_pomp(b_valc_ach_mean, 1, 7),
    values_hedonism_cons_b = convert_to_pomp(b_valc_hed_mean, 1, 7)
  ) %>%
  select(id, starts_with("purpose_"), starts_with("happiness_"), 
         starts_with("hope_"), starts_with("self_control"), 
         starts_with("autonomy_"), starts_with("competence_"), starts_with("relatedness_"),
         starts_with("distress_"), starts_with("experiential_"), starts_with("stress_"),
         starts_with("values_"))

# Create baseline correlations table (most complete data)
baseline_vars <- correlation_dat %>%
  select(
    # Purpose measures
    purpose_bpurp_b, purpose_pilea_total_b, purpose_pilea_eng_b, purpose_pilea_bey_b,
    # Happiness
    happiness_b,
    # Correlates
    hope_total_b, hope_pathways_b, hope_agency_b, self_control_b,
    autonomy_b, competence_b, relatedness_b,
    distress_intolerance_b, experiential_avoidance_b, stress_tolerance_b,
    values_benevolence_imp_b, values_universalism_imp_b, 
    values_achievement_imp_b, values_hedonism_imp_b,
    values_benevolence_cons_b, values_universalism_cons_b,
    values_achievement_cons_b, values_hedonism_cons_b
  )




```


# Results

## Correlations Between Purpose, Happiness, and Theoretically Relevant Measures

```{r}
#| label: correlation-analysis
#| echo: false
#| message: false
#| warning: false
#| include: true
#| results: 'asis'

#source('purpose_happiness_correlations.r')

cor_matrix <- cor(baseline_vars, use = "pairwise.complete.obs")

# Extract correlations with purpose and happiness measures
purpose_happiness_cors <- cor_matrix[6:23, 1:5] # Correlates with purpose/happiness vars

# Create a formatted correlation table
create_correlation_table <- function(cor_matrix, title = "Correlations") {
  cor_df <- as.data.frame(cor_matrix) %>%
    rownames_to_column("Variable") %>%
    mutate(
      Variable = case_when(
        Variable == "hope_total_b" ~ "Hope (Total)",
        Variable == "hope_pathways_b" ~ "Hope (Pathways)", 
        Variable == "hope_agency_b" ~ "Hope (Agency)",
        Variable == "self_control_b" ~ "Self-Control",
        Variable == "autonomy_b" ~ "Autonomy Need",
        Variable == "competence_b" ~ "Competence Need",
        Variable == "relatedness_b" ~ "Relatedness Need",
        Variable == "distress_intolerance_b" ~ "Distress Intolerance",
        Variable == "experiential_avoidance_b" ~ "Experiential Avoidance",
        Variable == "stress_tolerance_b" ~ "Stress Tolerance",
        Variable == "values_benevolence_imp_b" ~ "Benevolence Values (Importance)",
        Variable == "values_universalism_imp_b" ~ "Universalism Values (Importance)",
        Variable == "values_achievement_imp_b" ~ "Achievement Values (Importance)",
        Variable == "values_hedonism_imp_b" ~ "Hedonism Values (Importance)",
        Variable == "values_benevolence_cons_b" ~ "Benevolence Values (Consistency)",
        Variable == "values_universalism_cons_b" ~ "Universalism Values (Consistency)",
        Variable == "values_achievement_cons_b" ~ "Achievement Values (Consistency)",
        Variable == "values_hedonism_cons_b" ~ "Hedonism Values (Consistency)",
        TRUE ~ Variable
      )
    )
  
  # Rename columns
  names(cor_df)[2:6] <- c("Purpose (BPIL)", "Purpose (PILEA)", "Purpose (Engagement)", 
                          "Purpose (Beyond-Self)", "Happiness (SHS)")
  
  # Format to 3 decimal places
  cor_df[, 2:6] <- lapply(cor_df[, 2:6], function(x) sprintf("%.3f", x))
  
  return(cor_df)
}

# Create the correlation table
correlation_table <- create_correlation_table(purpose_happiness_cors)

# Display the table using kable
#print("Table 1: Correlations Between Purpose, Happiness, and Theoretically Relevant Measures")
#knitr::kable(correlation_table, 
#             caption = "Correlations Between Purpose, Happiness, and Theoretically Relevant Measures (Baseline)",
#             booktabs = TRUE) %>%
#  kable_styling(latex_options = c("striped", "hold_position")) %>%
#  group_rows("Time Orientation", 1, 4) %>%
#  group_rows("Psychological Needs", 5, 7) %>%
#  group_rows("Emotional Tolerance", 8, 10) %>%
#  group_rows("Values (Importance)", 11, 14) %>%
#  group_rows("Values (Behavioral Consistency)", 15, 18)

# Alternative GT table formatting (more publication-ready)
gt_table <- correlation_table %>%
  gt() %>%
  tab_header(
    title = "Correlations Between Purpose, Happiness, and Theoretically Relevant Measures",
    subtitle = "Baseline Assessment (N = 345)"
  ) %>%
  tab_row_group(
    label = "Time Orientation",
    rows = 1:4
  ) %>%
  tab_row_group(
    label = "Psychological Needs", 
    rows = 5:7
  ) %>%
  tab_row_group(
    label = "Emotional Tolerance",
    rows = 8:10
  ) %>%
  tab_row_group(
    label = "Values (Importance)",
    rows = 11:14
  ) %>%
  tab_row_group(
    label = "Values (Behavioral Consistency)",
    rows = 15:18
  ) %>%
  fmt_number(
    columns = 2:6,
    decimals = 3
  ) %>%
  cols_label(
    Variable = "Measure"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  tab_footnote(
    footnote = "All correlations based on POMP-standardized scores (0-100 scale). BPIL = Brief Purpose in Life Scale; PILEA = Purpose in Life External Aim Scale; SHS = Subjective Happiness Scale.",
    locations = cells_title()
  )

print(gt_table)
```

## Differential Correlation Analysis

```{r}
#| label: differential-correlation-analysis
#| echo: false
#| message: false
#| warning: false
#| include: true
#| results: 'asis'

# Create differential correlation analysis
# Test for significant differences using Fisher's z-transformation
test_correlation_difference <- function(r1, r2, n) {
  # Fisher z-transformation
  z1 <- 0.5 * log((1 + r1) / (1 - r1))
  z2 <- 0.5 * log((1 + r2) / (1 - r2))
  
  # Standard error
  se <- sqrt(1/(n-3) + 1/(n-3))
  
  # Z-test statistic
  z_stat <- (z1 - z2) / se
  
  # Two-tailed p-value
  p_value <- 2 * (1 - pnorm(abs(z_stat)))
  
  return(list(z = z_stat, p = p_value))
}

# Calculate sample size for complete cases at baseline
n_complete <- sum(complete.cases(baseline_vars))
cat("Sample size for correlation analysis:", n_complete, "\n")

# Test key theoretical differences
purpose_bpil_cors <- as.numeric(correlation_table$`Purpose (BPIL)`)
happiness_cors <- as.numeric(correlation_table$`Happiness (SHS)`)

# Test differences for key theoretical predictions
key_tests <- data.frame(
  Measure = correlation_table$Variable,
  Purpose_r = purpose_bpil_cors,
  Happiness_r = happiness_cors,
  Difference = purpose_bpil_cors - happiness_cors
)

# Add significance tests
key_tests$z_stat <- mapply(function(r1, r2) {
  test_correlation_difference(r1, r2, n_complete)$z
}, key_tests$Purpose_r, key_tests$Happiness_r)

key_tests$p_value <- mapply(function(r1, r2) {
  test_correlation_difference(r1, r2, n_complete)$p  
}, key_tests$Purpose_r, key_tests$Happiness_r)

# Add significance indicators
key_tests$Significance <- case_when(
  key_tests$p_value < 0.001 ~ "***",
  key_tests$p_value < 0.01 ~ "**", 
  key_tests$p_value < 0.05 ~ "*",
  TRUE ~ ""
)

# Format for publication
key_tests_formatted <- key_tests %>%
  mutate(
    Purpose_r = sprintf("%.3f", Purpose_r),
    Happiness_r = sprintf("%.3f", Happiness_r), 
    Difference = sprintf("%.3f", Difference),
    z_stat = sprintf("%.2f", z_stat),
    p_value = ifelse(p_value < 0.001, "< .001", sprintf("%.3f", p_value))
  ) %>%
  select(Measure, Purpose_r, Happiness_r, Difference, z_stat, p_value, Significance)

# Create publication table for differential correlations
differential_table <- key_tests_formatted %>%
  gt() %>%
  tab_header(
    title = "Differential Correlations: Purpose vs. Happiness",
    subtitle = "Statistical Tests of Correlation Differences (Baseline)"
  ) %>%
  cols_label(
    Measure = "Measure",
    Purpose_r = "Purpose (BPIL)",
    Happiness_r = "Happiness (SHS)", 
    Difference = "Difference",
    z_stat = "z-statistic",
    p_value = "p-value",
    Significance = ""
  ) %>%
  tab_row_group(
    label = "Time Orientation",
    rows = 1:4
  ) %>%
  tab_row_group(
    label = "Psychological Needs", 
    rows = 5:7
  ) %>%
  tab_row_group(
    label = "Emotional Tolerance",
    rows = 8:10
  ) %>%
  tab_row_group(
    label = "Values (Importance)",
    rows = 11:14
  ) %>%
  tab_row_group(
    label = "Values (Behavioral Consistency)",
    rows = 15:18
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  ) %>%
  tab_style(
    style = cell_fill(color = "lightblue", alpha = 0.3),
    locations = cells_body(
      columns = Significance,
      rows = Significance %in% c("*", "**", "***")
    )
  ) %>%
  tab_footnote(
    footnote = "* p < .05, ** p < .01, *** p < .001. Positive differences indicate stronger correlation with purpose than happiness.",
    locations = cells_title()
  )

print(differential_table)
```

## Longitudinal Stability Analysis

```{r}
#| label: longitudinal-stability-analysis
#| echo: false
#| message: false
#| warning: false
#| include: true
#| results: 'asis'

# Longitudinal Stability Analysis
longitudinal_vars <- correlation_dat %>%
  select(
    # Purpose measures across time
    purpose_bpurp_b, purpose_bpurp_fu1, purpose_bpurp_fu2,
    purpose_pilea_total_b, purpose_pilea_total_fu1, purpose_pilea_total_fu2,
    
    # Happiness across time  
    happiness_b, happiness_fu1, happiness_fu2,
    
    # Key correlates across time (where available)
    hope_total_b, hope_total_fu1, hope_total_fu2,
    autonomy_b, autonomy_fu1, autonomy_fu2,
    competence_b, competence_fu1, competence_fu2,
    relatedness_b, relatedness_fu1, relatedness_fu2,
    stress_tolerance_b, stress_tolerance_fu1, stress_tolerance_fu2
  )

# Calculate stability correlations
stability_cors <- cor(longitudinal_vars, use = "pairwise.complete.obs")

# Extract key stability correlations
purpose_stability <- data.frame(
  Measure = c("Purpose (BPIL)", "Purpose (PILEA)", "Happiness", 
              "Hope", "Autonomy", "Competence", "Relatedness", "Stress Tolerance"),
  `T1-T2` = c(
    stability_cors["purpose_bpurp_b", "purpose_bpurp_fu1"],
    stability_cors["purpose_pilea_total_b", "purpose_pilea_total_fu1"], 
    stability_cors["happiness_b", "happiness_fu1"],
    stability_cors["hope_total_b", "hope_total_fu1"],
    stability_cors["autonomy_b", "autonomy_fu1"],
    stability_cors["competence_b", "competence_fu1"], 
    stability_cors["relatedness_b", "relatedness_fu1"],
    stability_cors["stress_tolerance_b", "stress_tolerance_fu1"]
  ),
  `T1-T3` = c(
    stability_cors["purpose_bpurp_b", "purpose_bpurp_fu2"],
    stability_cors["purpose_pilea_total_b", "purpose_pilea_total_fu2"],
    stability_cors["happiness_b", "happiness_fu2"], 
    stability_cors["hope_total_b", "hope_total_fu2"],
    stability_cors["autonomy_b", "autonomy_fu2"],
    stability_cors["competence_b", "competence_fu2"],
    stability_cors["relatedness_b", "relatedness_fu2"],
    stability_cors["stress_tolerance_b", "stress_tolerance_fu2"]
  ),
  `T2-T3` = c(
    stability_cors["purpose_bpurp_fu1", "purpose_bpurp_fu2"],
    stability_cors["purpose_pilea_total_fu1", "purpose_pilea_total_fu2"],
    stability_cors["happiness_fu1", "happiness_fu2"],
    stability_cors["hope_total_fu1", "hope_total_fu2"],
    stability_cors["autonomy_fu1", "autonomy_fu2"],
    stability_cors["competence_fu1", "competence_fu2"],
    stability_cors["relatedness_fu1", "relatedness_fu2"],
    stability_cors["stress_tolerance_fu1", "stress_tolerance_fu2"]
  ),
  check.names = FALSE
)

# Format stability table
stability_table <- purpose_stability %>%
  mutate(
    across(c(`T1-T2`, `T1-T3`, `T2-T3`), ~ sprintf("%.3f", .))
  ) %>%
  gt() %>%
  tab_header(
    title = "Temporal Stability of Purpose, Happiness, and Key Correlates",
    subtitle = "Test-Retest Correlations Across Three Time Points"
  ) %>%
  cols_label(
    Measure = "Measure",
    `T1-T2` = "Baseline - 6 months",
    `T1-T3` = "Baseline - 2 years", 
    `T2-T3` = "6 months - 2 years"
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body(rows = 1:3)
  ) %>%
  tab_footnote(
    footnote = "T1 = Baseline, T2 = 6-month follow-up, T3 = 2-year follow-up. Bold rows indicate primary constructs of interest.",
    locations = cells_title()
  )

print(stability_table)
```

## Cross-Lagged Associations Between Purpose and Happiness

```{r}
#| label: cross-lagged-analysis
#| echo: false
#| message: false
#| warning: false
#| include: true
#| results: 'asis'


# Cross-lagged correlations
cross_lagged <- data.frame(
  Predictor = c("Baseline Purpose (BPIL)", "Baseline Purpose (PILEA)", "Baseline Happiness",
               "Baseline Purpose (BPIL)", "Baseline Purpose (PILEA)", "Baseline Happiness"),
  Outcome = c("6-month Happiness", "6-month Happiness", "6-month Purpose (BPIL)",
             "2-year Happiness", "2-year Happiness", "2-year Purpose (BPIL)"),
  Correlation = c(
    stability_cors["purpose_bpurp_b", "happiness_fu1"],
    stability_cors["purpose_pilea_total_b", "happiness_fu1"],
    stability_cors["happiness_b", "purpose_bpurp_fu1"],
    stability_cors["purpose_bpurp_b", "happiness_fu2"],
    stability_cors["purpose_pilea_total_b", "happiness_fu2"], 
    stability_cors["happiness_b", "purpose_bpurp_fu2"]
  )
) %>%
  mutate(
    Correlation = sprintf("%.3f", Correlation),
    Time_Frame = rep(c("6-month", "2-year"), each = 3)
  )

# Format cross-lagged table
cross_lagged_table <- cross_lagged %>%
  gt() %>%
  tab_header(
    title = "Cross-Lagged Associations Between Purpose and Happiness", 
    subtitle = "Longitudinal Predictive Relationships"
  ) %>%
  tab_row_group(
    label = "6-Month Predictions",
    rows = 1:3
  ) %>%
  tab_row_group(
    label = "2-Year Predictions", 
    rows = 4:6
  ) %>%
  cols_hide(Time_Frame) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels()
  ) %>%
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_row_groups()
  )

print(cross_lagged_table)

# Summary of key findings for manuscript
#cat("\n=== KEY FINDINGS SUMMARY ===\n")
#cat("1. Correlation Magnitudes:\n")
#cat("   - Purpose-Happiness correlation (baseline):", sprintf("%.3f", cor(correlation_dat$purpose_bpurp_b, correlation_dat$happiness_b, use = "complete.obs")), "\n")
#cat("   - Purpose (BPIL) stability (baseline to 2-year):", sprintf("%.3f", stability_cors["purpose_bpurp_b", "purpose_bpurp_fu2"]), "\n")
#cat("   - Happiness stability (baseline to 2-year):", sprintf("%.3f", stability_cors["happiness_b", "happiness_fu2"]), "\n")
#cat("\n2. Differential Correlations (Purpose vs Happiness):\n")
```










# Discussion

The results from this comprehensive longitudinal analysis reveal important distinctions between purpose in life and happiness. Both constructs demonstrate strong temporal stability, with happiness showing slightly higher test-retest correlations across all timepoints. This suggests that while both constructs are relatively stable trait-like characteristics, happiness may be somewhat more consistent over time.

## Key Findings

